---
title: "Gapminder Explorer"
output: html_document
---
## Overview
* #### *Bringing in the data*
* #### *Smell testing data*
* #### *Exploring individual variables*
* #### *Exploring various plot types*
* #### *Doing more!*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Bringing in data
First lets begin by loading the gapminder dataset and the tidyverse package. Both must first be installed in the console using the `install.packages("")` function. 


Now the packages can be primed using the `library()` function for both `gapminder` and `tidyverse`


```{r,  echo=TRUE}
library(gapminder)
library(tidyverse)
```

## Smell testing data

### Structure of the gapminder object

The `str()` command can tell us about the structure of the gapminder object:

```{r, echo=TRUE}
str(gapminder)
```
##### Object

* A general look at the output shows us several variables and the first few values for each observation 
* This tells us this is 2-dimensional data with different variable types making this dataset a dataframe object


##### Class

* The output defines class of this dataset in the first line as a: tibble_dataframe, tibble and data.frame

##### Number of variables/columns

* There are 6 variables, meaning there are 6 columns

##### Number of observations/rows

* There are 1704 observations, meaning there are 1704 rows

##### Data type of each variable

* Country and continent variables are of the type factor
* Year and population are of the type integer
* Life expectancy is of the type numerical

### Summary of the gapminder object

Alternatively, the `summary()` method can perform some simple data transformations and provide us with a 6-digit summary for numerical variables (integer, numerical) or counts for categorical variables (factor). This would be more useful to get a simple look at the distribution of each variable.

```{r, echo=TRUE}
summary(gapminder)
```

## Exploring individual variable

#### Categorical variable: continent

First, select the categorical variable, "continent", and store in the value "c". Then use the `summary()` method to retrieve a summary of this variable. We can see there are 5 continents here: Africa, Americas, Asia, Europe and Oceania. The number of obserations with each continent are tallied alongside.  
```{r, echo=TRUE}
#first select all observations of the continent variable and place in the value 'c'

c <- select(gapminder, continent)

#use the summary() method to retrieve desired statistics

summary(c)

#next lets plot a bar graph of the counts of each continent

ggplot(c) + geom_bar(mapping=aes(x=continent))

```


#### Numerical variable: life expectancy

Next lets look at the life expectancy variable by storing it in the value "p" along with year (2007 only) and continent for further analysis. Using the `summary()` method again, we get a 6-digit summary (min,1st quartile, median, mean, 3rd quartile, max) describing the distribution of this variable. We can then plot the life expectancy according to continent in a boxplot. Clearly, Africa has the widest range of life expectancy and the lowest mean. Asia and America have notable outliers. Oceania has the highest avereage life expectancy.

```{r, echo=TRUE}
#place lifeExp variable of gapminder in "p", along with continent and year for further analysis

p <- select(gapminder, lifeExp,continent,year) %>% filter(year == 2007)

#summarize 'p' with 6-digit summary

summary(p)

#next lets look at the distribution of the population variable

ggplot(p) + geom_boxplot(mapping=aes(x=continent, y = lifeExp))
```

## Exploring various plot types

Here we're going to compare the life expectancy distribution of Asia and Europe in the year 2007. First, you `filter` out the gapminder dataset for observations that have either "Asia" or "Europe" as their continent and year as 2007. Then you pipe this into the `select()` method to select only life expectancy and continent for these observations. Then we plot a kernal density plot of life expectancy where each line represents a continent. This plot shows that Europe has, on average, a higher life expectancy than Asia. There is also a narrower distribution for Europe that ranges from about 67-83 years whereas the wider distribution of Asia ranges from 34-83. Europe also has a notable binomial peak, with one peak around 80 years and one around 75 years. 

```{r, echo=TRUE}
#use of filter() and select() to place life expectancy of observations that are only from Asia or Europe in the variable 'g'

g <- filter(gapminder, continent %in% c("Asia", "Europe"), year == 2007) %>% select(lifeExp, continent)

#plotting kernal density plot of life expectancy with each line representing continent

ggplot(g) + geom_density(mapping=aes(x=lifeExp, linetype = continent))
```

Next lets look at life expectancy in relation to GDP per capita. First we filter out the observations for the year 2007, then creating a dot plot with a line of best fit of life expectancy on the y-axis and gdpPercap on the x-axis. This graph shows us that there is a fair connection between the two variables as there is little variability among observations. At high gdpPercap, life expectancy remains relatively high despite changes in gdpPercap until a certain threshold is reached, below of which there is a signficant reduction in life expectancy in relation to a drop in gdpPercap.

```{r, echo=TRUE}
#filter observations for only those from the year 2007

k <- filter(gapminder, year == 2007)

#plot the gdpPercap versus the life expectancy and incorporate a line of best fit

ggplot(k, mapping = aes(x=gdpPercap, y=lifeExp)) + geom_point() + geom_smooth()
```

## Doing more!

Here is a way to filter information according to a vector of possible values. We want to collect observations who's country is either Rwanda or Afghanistan so we can use the `filter()` method and set country to the vector `c("Rwanda", "Afghanistan")`

```{r, echo=TRUE}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```












