---
title: "Gapminder-Explorer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Bringing in data
First lets begin by loading the gapminder dataset and the tidyverse package. Both must first be installed in the console using the `install.packages("")` function. 

Now the packages can be primed using the `library()` function for both `"gapminder"` and `"tidyverse"`


```{r, include=FALSE}
library(gapminder)
library(tidyverse)
```

## Smell testing data

### Structure of the gapminder object

The `str()` command can tell us about the structure of the gapminder object:

```{r}
str(gapminder)
```
##### Object

* A general look at the output shows us several variables and the first few values for each observation which tells us this is 2-dimensional data with different variable types making this a dataframe


##### Class

* The output defines class in the first line as a: tibble_dataframe, tibble and data.frame

##### Number of variables/columns

* There are 6 variables, meaning there are 6 columns

##### Number of observations/rows

* There are 1704 observations, meaning there are 1704 rows

##### Data type of each variable

* Country and continent variables are of the type factor
* Year and population are of the type integer
* Life expectancy is of the type numerical

### Summary of the gapminder object

Alternatively, the `summary()` method can perform some simple data transformations and provide us with a 6-digit summary for numerical variables (integer, numerical) or counts for categorical variables (factor). This would be more useful to get a simple look at the distribution of each variable.

```{r}
summary(gapminder)
```

## Exploring individual variable

#### Categorical variable: continent

First, select the categorical variable, "continent", and store in the value "c". Then use the `summary()` method to retrieve a summary of this variable. We can see there are 5 continents here: Africa, Americas, Asia, Europe and Oceania. The number of obserations with each continent are tallied alongside.  
```{r}
#first select all observations of the continent variable and place in the value 'c'
c <- select(gapminder, continent)

#use the summary() method to retrieve desired statistics
summary(c)

#ggplot(c)+ geom_area(mapping=aes(stat="bin"))
```


#### Numerical variable: population

Next lets look at the population variable by storing it in the value "p". Using the `summary()` method again, we get a 6-digit summary (min,) describing the distribution of this variable

```{r}
#place pop variable of gapminder in "p"
p <- select(gapminder, pop)

#summarize 'p' with 6-digit summary
summary(p)
```

## Exploring various plot types

Here we're going to compare the life expectancy distribution of Asia and Europe in the year 2007. First, you `filter` out the gapminder dataset for observations that have either "Asia" or "Europe" as their continent and year as 2007. Then you pipe this into the `select()` method to select only life expectancy and continent for these observations. Then we plot a kernal density plot of life expectancy where each line represents a continent. This plot shows that Europe has, on average, a higher life expectancy than Asia. There is also a narrower distribution for Europe that ranges from about 67-83 years whereas the wider distribution of Asia ranges from 34-83. Europe also has a notable binomial peak, with one peak around 80 years and one around 75 years. 

```{r}
#use of filter() and select() to place life expectancy of observations that are only from Asia or Europe in the variable 'g'
g <- filter(gapminder, continent %in% c("Asia", "Europe"), year == 2007) %>% select(lifeExp, continent)

#plotting kernal density plot of life expectancy with each line representing continent
ggplot(g) + geom_density(mapping=aes(x=lifeExp, linetype = continent))
```

Next lets look at life expectancy in relation to GDP per capita. First we filter out the observations for the year 2007, then creating a dot plot with a line of best fit of life expectancy on the y-axis and gdpPercap on the x-axis. This graph shows us that there is a fair connection between the two variables as there is little variability among observations. At high gdpPercap, life expectancy remains relatively high despite changes in gdpPercap until a certain threshold is reached, below of which there is a signficant reduction in life expectancy in relation to a drop in gdpPercap.

```{r}
#filter observations for only those from the year 2007
k <- filter(gapminder, year == 2007)

#plot the gdpPercap versus the life expectancy and incorporate a line of best fit
ggplot(k, mapping = aes(x=gdpPercap, y=lifeExp)) + geom_point() + geom_smooth()
```

## Doing more!

Here is a way to filter information according to a vector of possible values. We want to collect observations who's country is either Rwanda or Afghanistan so we can use the `filter()` method and set country to the vector `c("Rwanda", "Afghanistan")`

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```












